\large{\bf Looking forwards and backwards: dynamics and genealogies of locally regulated populations }

### 2023-12-11 forwards-and-backwards.tex
alison fixup to PME proof

We shall next show that $\zeta^\epsilon * \psi_t^\epsilon$ solves \eqref{PME1}
up to a remainder of order $\epsilon$.
First observe that
\begin{equation}
\label{total equation}
\int\Delta\left((\rho^{\epsilon}*\psi_t^\epsilon)\psi_t^\epsilon\right)\phi  dx
    =
    - \int \nabla(\rho^{\epsilon}*\psi_t^\epsilon)\, \psi_t^\epsilon\,\nabla\phi dx
    -\int \rho^{\epsilon}*\psi_t^\epsilon \,\nabla \psi_t^\epsilon \, \nabla\phi dx.
\end{equation}
We would like to show that this is close to $\int (\zeta^\epsilon * \psi^\epsilon_t)^2 \Delta \phi dx$.
For the first term
\begin{eqnarray}
\nonumber
\int \nabla(\rho^{\epsilon}*\psi_t^\epsilon)\, \psi_t^\epsilon\, \nabla\phi  dx
&=&
\int\int\int\nabla \psi_t^\epsilon(x-y)\zeta^{\epsilon}(y-z)
\check{\zeta}^{\epsilon}(z)
\psi_t^\epsilon(x)\nabla\phi(x) dz  dy  dx
\\
\nonumber
&=&
\int\int\int\nabla \psi_t^\epsilon(\tilde{x}-\tilde{y})
\zeta^{\epsilon}(\tilde{y})
\zeta^{\epsilon}(\tilde{z})\psi_t^\epsilon(\tilde{x}-\tilde{z})
\nabla\phi(\tilde{x}-\tilde{z}) d\tilde{z} d\tilde{y} d\tilde{x}
\\
\nonumber
&=&\int(\nabla \zeta^{\epsilon}*\psi_t^\epsilon)\,
\left(\zeta^{\epsilon}*(\psi_t^\epsilon\nabla\phi)\right)
d x\\
\nonumber
&=&
\frac{1}{2}\int\nabla ((\zeta^{\epsilon}*\psi_t^\epsilon)^2)
\,\nabla\phi  dx
\\
\label{error 1}
&&+
\int\nabla(\zeta^{\epsilon}*\psi_t^\epsilon)
\left[\zeta^{\epsilon}*(\psi_t^\epsilon\,\nabla\phi)-
\nabla\phi\,(\zeta^{\epsilon}*\psi_t^\epsilon)\right] dx,
\end{eqnarray}
where, as before, we have substituted $\tilde{x}=x-z$, $\tilde{y}=y-z$,
$\tilde{z}=-z$.
To control the term~(\ref{error 1})
we use the Intermediate Value Theorem to see that
\begin{multline*}
\left|\int\left[\psi_t^\epsilon(x-y)\nabla\phi(x-y)\zeta^{\epsilon}(y)-
\nabla\phi(x)\psi_t^\epsilon(x-y)\zeta^{\epsilon}(y)
\right] dy\right|
\\
\leq C\|\Delta\phi\|_\infty\int \psi_t^\epsilon (x-y)y\zeta^{\epsilon}(y) d y.
\end{multline*}
Since $\zeta\in\mathcal{S}(\IR)$, 
the integral in this expression is
$\mathcal{O}(\epsilon)$.  \revpoint{1}{31}

Similarly, for the second term in~(\ref{total equation}),
\begin{eqnarray}
\nonumber
\int (\rho^{\epsilon}*\psi_t^\epsilon)\, \nabla \psi_t^\epsilon\, \nabla\phi dx
&=&
\int\int\int \psi_t^\epsilon(x-y)\zeta^{\epsilon}(y-z)
\check{\zeta}^{\epsilon}(z)\nabla \psi_t^\epsilon(x)\nabla\phi(x) dz dy  dx
\\
\nonumber
&=&
\int\int\int \psi_t^\epsilon(\tilde{x}-\tilde{y})\zeta^{\epsilon}(\tilde{y})
\zeta^{\epsilon}(\tilde{z})\nabla \psi_t^\epsilon(\tilde{x}-\tilde{z})
\nabla\phi(\tilde{x}-\tilde{z}) d \tilde{z} d\tilde{y} d\tilde{x}
\\
\nonumber
	&=&\int(\zeta^{\epsilon}*\psi_t^\epsilon)\,
\left(\zeta^{\epsilon}*(\nabla \psi_t^\epsilon \, \nabla\phi)\right)  dx\\
\nonumber
&=&
\frac{1}{2}\int\nabla ((\zeta^{\epsilon}*\psi_t^\epsilon)^2)
\, \nabla\phi  dx
\\
\label{error 2}
&&+
\int (\zeta^{\epsilon}*\psi_t^\epsilon)
\left[\zeta^{\epsilon}*(\nabla \psi_t^\epsilon\, \nabla\phi)-
\nabla\phi\,(\nabla \zeta^{\epsilon}*\psi_t^\epsilon)\right] dx,
\end{eqnarray}
and~(\ref{error 2}) is controlled in the same way as~(\ref{error 1}):
using the Intermediate Value Theorem,
\begin{multline*}
\left|\int\left[\nabla \psi_t^\epsilon(x-y)\nabla\phi(x-y)\zeta^{\epsilon}(y)
-\nabla\phi(x)\nabla \psi_t^\epsilon(x-y)\zeta^{\epsilon}(y)
\right] dy\right|
\\
\leq C\|\Delta\phi\|_\infty\left|\int \nabla \psi_t^\epsilon(x-y)
y \zeta^{\epsilon}(y)
d y\right|,
\end{multline*}
which again is $\mathcal{O}(\epsilon)$.

We now have the ingredients that we need.
The calculations above yield both a uniform (in $\epsilon$) bound on
$\zeta^{\epsilon}*\psi_t^\epsilon$ in
$L^1\cap L^2\big([0,T]\times \IR\big)$, and that
\begin{multline} \label{eqn:weak_approx_pme}
\int \psi_t^\epsilon(x)\phi(x) dx-\int \psi_0^\epsilon (x)\phi(x) dx
	=\int_0^t\int (\zeta^{\epsilon}*\psi_s^\epsilon(x))^2\Delta\phi(x) dx
\\
	+\int_0^t \int \zeta^{\epsilon}*\psi_s^\epsilon(x)
	\left(1- \zeta^{\epsilon}*\psi_s^\epsilon(x)\right)
\phi(x) dx
+\mathcal{O}(\epsilon)
\end{multline}
(for sufficiently regular $\phi$).
Since
$\int \psi_t^\epsilon(x)\phi(x) d x- \int \zeta^{\epsilon}*\psi_t^\epsilon (x)\phi(x) d x$
is order $\epsilon$,
if we replace $\psi^\epsilon$ by $\zeta^{\epsilon}*\psi^\epsilon$ on the left hand side,
then~\eqref{eqn:weak_approx_pme} says that $\zeta^{\epsilon}*\psi^\epsilon$
solves~\eqref{PME} weakly up to order $\epsilon$.
Therefore, $\zeta^{\epsilon}*\psi^\epsilon$ converges weakly to $\psi$ in $L^1$,
where $\psi$ is the (unique)
solution to equation~(\ref{PME})
and, so, therefore, does $\psi^\epsilon$. In fact, strong convergence, that is
$\int |\psi^\epsilon -\psi|\phi dx \to 0$, follows from the
uniform integrability of $\psi^\epsilon$ that we
can deduce from the uniform control of
$\int \psi^\epsilon|\log \psi^\epsilon| d x$ that we proved above.
\end{proof}

>>>>

We shall next show that $\zeta^\epsilon * \psi_t^\epsilon$ solves \eqref{PME1}
up to a remainder of order $\epsilon$.
First observe that
\begin{equation}
\label{total equation}
\int\Delta\left((\rho^{\epsilon}*\psi_t^\epsilon)\psi_t^\epsilon\right)\phi  dx
    =
    - \int \nabla(\rho^{\epsilon}*\psi_t^\epsilon)\, \psi_t^\epsilon\,\nabla\phi dx
    -\int \rho^{\epsilon}*\psi_t^\epsilon \,\nabla \psi_t^\epsilon \, \nabla\phi dx.
\end{equation}
We would like to show that this is close to $\int (\zeta^\epsilon * \psi^\epsilon_t)^2 \Delta \phi dx$.
For the first term
\begin{eqnarray}
\nonumber
\int \nabla(\rho^{\epsilon}*\psi_t^\epsilon)\, \psi_t^\epsilon\, \nabla\phi  dx
&=&
\int\int\int\nabla \psi_t^\epsilon(x-y)\zeta^{\epsilon}(y-z)
\check{\zeta}^{\epsilon}(z)
\psi_t^\epsilon(x)\nabla\phi(x) dz  dy  dx
\\
\nonumber
&=&
\int\int\int\nabla \psi_t^\epsilon(\tilde{x}-\tilde{y})
\zeta^{\epsilon}(\tilde{y})
\zeta^{\epsilon}(\tilde{z})\psi_t^\epsilon(\tilde{x}-\tilde{z})
\nabla\phi(\tilde{x}-\tilde{z}) d\tilde{z} d\tilde{y} d\tilde{x}
\\
\nonumber
&=&\int(\nabla \zeta^{\epsilon}*\psi_t^\epsilon)\,
\left(\zeta^{\epsilon}*(\psi_t^\epsilon\nabla\phi)\right)
d x\\
\nonumber
&=&
\frac{1}{2}\int\nabla ((\zeta^{\epsilon}*\psi_t^\epsilon)^2)
\,\nabla\phi  dx
\\
\label{error 1}
&&+
\int\nabla(\zeta^{\epsilon}*\psi_t^\epsilon)
\left[\zeta^{\epsilon}*(\psi_t^\epsilon\,\nabla\phi)-
\nabla\phi\,(\zeta^{\epsilon}*\psi_t^\epsilon)\right] dx,
\end{eqnarray}
where, as before, we have substituted $\tilde{x}=x-z$, $\tilde{y}=y-z$,
$\tilde{z}=-z$.
We are going to bound the square of the $L^2$-norm of the term in square brackets 
in~(\ref{error 1}) by the product of its $L^1$-norm (which is bounded
by a constant times the $L^1$-norm of $\psi_t^\epsilon$) and its $L^\infty$-norm.
To control the $L^\infty$-norm,
we use the Intermediate Value Theorem to see that
\begin{multline*}
\left|\int\left[\psi_t^\epsilon(x-y)\nabla\phi(x-y)\zeta^{\epsilon}(y)-
\nabla\phi(x)\psi_t^\epsilon(x-y)\zeta^{\epsilon}(y)
\right] dy\right|
\\
	\leq C\|\Delta\phi\|_\infty\int \psi_t^\epsilon (x-y)
	\epsilon\frac{|y|}{\epsilon}\zeta^{\epsilon}(y) d y
	\leq C\|\Delta\phi\|_\infty\epsilon \|z\zeta(z)\|_\infty\int \psi_t^\epsilon (x-y) dy.
	%|y|\zeta^{\epsilon}(y) d y
\end{multline*}
Since $\zeta\in\mathcal{S}(\IR)$, and as we checked above $\psi_t^\epsilon$ is 
uniformly bounded in $L^1$ over $[0,T]$,
this expression is
$\mathcal{O}(\epsilon)$.  \revpoint{1}{31}

We can now apply the Cauchy-Schwarz inequality to~(\ref{error 1}) to bound
it by the square root of
\begin{equation}
	C\epsilon\int \big(\nabla(\zeta^\epsilon*\psi_t^\epsilon(x))\big)^2 dx.
\end{equation}

Similarly, for the second term in~(\ref{total equation}),
\begin{eqnarray}
\nonumber
\int (\rho^{\epsilon}*\psi_t^\epsilon)\, \nabla \psi_t^\epsilon\, \nabla\phi dx
&=&
\int\int\int \psi_t^\epsilon(x-y)\zeta^{\epsilon}(y-z)
\check{\zeta}^{\epsilon}(z)\nabla \psi_t^\epsilon(x)\nabla\phi(x) dz dy  dx
\\
\nonumber
&=&
\int\int\int \psi_t^\epsilon(\tilde{x}-\tilde{y})\zeta^{\epsilon}(\tilde{y})
\zeta^{\epsilon}(\tilde{z})\nabla \psi_t^\epsilon(\tilde{x}-\tilde{z})
\nabla\phi(\tilde{x}-\tilde{z}) d \tilde{z} d\tilde{y} d\tilde{x}
\\
\nonumber
	&=&\int(\zeta^{\epsilon}*\psi_t^\epsilon)\,
\left(\zeta^{\epsilon}*(\nabla \psi_t^\epsilon \, \nabla\phi)\right)  dx\\
\nonumber
&=&
\frac{1}{2}\int\nabla ((\zeta^{\epsilon}*\psi_t^\epsilon)^2)
\, \nabla\phi  dx
\\
\label{error 2}
&&+
\int (\zeta^{\epsilon}*\psi_t^\epsilon)
\left[\zeta^{\epsilon}*(\nabla \psi_t^\epsilon\, \nabla\phi)-
\nabla\phi\,(\nabla \zeta^{\epsilon}*\psi_t^\epsilon)\right] dx.
\end{eqnarray}
The term in square brackets in~(\ref{error 2}) is %controlled in the same way as~(\ref{error 1}):
%using the Intermediate Value Theorem,
\begin{multline*}
\left|\int\left[\nabla \psi_t^\epsilon(x-y)\nabla\phi(x-y)\zeta^{\epsilon}(y)
-\nabla\phi(x)\nabla \psi_t^\epsilon(x-y)\zeta^{\epsilon}(y)
\right] dy\right|
\\
=
	\left|\int\left[\psi_t^\epsilon(x-y)
	\nabla\big((\nabla\phi(x-y)-\nabla\phi(x))\zeta^{\epsilon}(y)\big)
%-\nabla\phi(x)\nabla \psi_t^\epsilon(x-y)\zeta^{\epsilon}(y)
\right] dy\right|,
%\leq C\|\Delta\phi\|_\infty\left|\int \nabla \psi_t^\epsilon(x-y)
%y \zeta^{\epsilon}(y)
%d y\right|,
\end{multline*}
and expanding $\nabla\phi$ in a Taylor series about $x$ and once again using that
$\zeta\in \mathcal{S}(\IR)$, we see that the $L^\infty$ norm of this quantity is
$\mathcal{O}(\epsilon)$.

We can now apply the Cauchy-Schwarz inequality to~(\ref{error 2}) to bound
it by the square root of
\begin{equation}
	C\epsilon\int \big(\zeta^\epsilon*\psi_t^\epsilon(x)\big)^2 dx.
\end{equation}

We now have the ingredients that we need. Recalling that 
$\int_0^t \int(\zeta^{\epsilon}*\psi_s^\epsilon(x))^2  dx ds$ and
$\int_0^t \int \big(\nabla (\zeta^{\epsilon}*\psi_s^\epsilon(x))\big)^2  dx ds$
are uniformly bounded in
$\epsilon$ and $t\in [0,T]$,
the calculations above yield both a uniform (in $\epsilon$) bound on
$\zeta^{\epsilon}*\psi_t^\epsilon$ in
$L^1\cap L^2\big([0,T]\times \IR\big)$, and (with another application of 
Cauchy-Schwarz, this time applied to the time integral, to control the error terms) that
\begin{multline} \label{eqn:weak_approx_pme}
\int \psi_t^\epsilon(x)\phi(x) dx-\int \psi_0^\epsilon (x)\phi(x) dx
	=\int_0^t\int (\zeta^{\epsilon}*\psi_s^\epsilon(x))^2\Delta\phi(x) dx
\\
	+\int_0^t \int \zeta^{\epsilon}*\psi_s^\epsilon(x)
	\left(1- \zeta^{\epsilon}*\psi_s^\epsilon(x)\right)
\phi(x) dx
	+\mathcal{O}(\sqrt{\epsilon})
\end{multline}
(for sufficiently regular $\phi$).
Since
$\int \psi_t^\epsilon(x)\phi(x) d x- \int \zeta^{\epsilon}*\psi_t^\epsilon (x)\phi(x) d x$
is order $\epsilon$,
if we replace $\psi^\epsilon$ by $\zeta^{\epsilon}*\psi^\epsilon$ on the left hand side,
then~\eqref{eqn:weak_approx_pme} says that $\zeta^{\epsilon}*\psi^\epsilon$
solves~\eqref{PME} weakly up to order $\epsilon$.
Therefore, $\zeta^{\epsilon}*\psi^\epsilon$ converges weakly to $\psi$ in $L^1$,
where $\psi$ is the (unique)
solution to equation~(\ref{PME})
and, so, therefore, does $\psi^\epsilon$. In fact, strong convergence, that is
$\int |\psi^\epsilon -\psi|\phi dx \to 0$, follows from the
uniform integrability of $\psi^\epsilon$ that we
can deduce from the uniform control of
$\int \psi^\epsilon|\log \psi^\epsilon| d x$ that we proved above.
\end{proof}
